---
title: "Exploring Income/Poverty in Colorado Using R"
author: "Nichole Monhait"
date: 2020-04-25
tags: ["R Markdown", "plot", "income"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

```{r, include = FALSE}
library(tidyverse)
library(readr)
```

### Data

The Colorado Department of Public Health and Environment (CDPHE) has an Open Data portal for many health indicators, linked [here](https://data-cdphe.opendata.arcgis.com/). Data can be downloaded in .csv format, which is what I did here for the Income/Poverty dataset. The purpose of this project is to provide an easy framework to generate a similar visualization. This project has no relation to the accuracy of the data nor the insight gathered.  I used the following code to bring the data into RStudio. 

```{r, message=FALSE, warning=FALSE}
co_income <- read_csv("/Users/monhait/Desktop/r/nichole/data/IncomePoverty_Census_Tracts.csv", col_names = TRUE)
```

### Data Preparation

I decided that my end product would be a visualization of Colorado counties with the highest percentage of people below the poverty limit. I arbitrarily decided that my cutoff would be at the top 10 counties, so set `n=10` at the beginning of my script.   

To prepare the data, I first selected the columns `County` and `Percent_Poverty_AllPeople_Income_Below_Pov_Level`. Luckily, this dataset was already fairly clean. I then grouped on `County`, to ultimately create the `average_poverty` column which found the average percentage of people below the poverty limit per county. If you take a look at the unprepared data, you will see that there are several rows for each county, which is why I had to group them. After, I made the `County` variable a factor and lumped to get my 10 columns. I used `n=n` within the `fct_lump` call to designate my cutoff of 10 that I discussed above. Using `fct_lump` allows you to do this and all other counties then get an `Other` designation, which was filtered out. Lastly, I added a `plot_text` column that will later be used in the ggplot call. 

```{r, message=FALSE, warning=FALSE}
n <- 10

poverty_by_county <- co_income %>% 
  select(County, Percent_Poverty_AllPeople_Income_Below_Pov_Level) %>% 
  group_by(County) %>% 
  summarize(average_poverty = mean(Percent_Poverty_AllPeople_Income_Below_Pov_Level, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(County = as_factor(str_to_title(County)) %>% fct_lump(n = n, w = average_poverty) %>% 
           fct_reorder(average_poverty)) %>% 
  mutate_if(is.numeric, ~round(., 1)) %>% 
  filter(County != "Other") %>%    
  mutate(plot_text = str_glue("County: {County}\nPercent Poverty: {average_poverty}%"))
```

To create the lollipop chart, I used `geom_segment` and `geom_point`. Additional customization code is shown below.  

```{r, message=FALSE, warning=FALSE}
co_county_poverty_plot <- poverty_by_county %>% 
  
  #Geometries
  ggplot(aes(County, average_poverty)) +
  geom_segment(aes(yend = 0, xend = County)) +
  geom_point() +
  geom_text(aes(label = plot_text),
            hjust = "outward",
            size = 3) +
  
  #Formatting
  scale_y_continuous(limits = c(0,50), expand = c(0,0)) +
  coord_flip() +

  #Labels
  labs(title = "Colorado Counties with the Highest Poverty Percentage",
       subtitle = str_glue("Top {n} based on CDPHE Open Data Portal"),
       y = "Avg. Percent of all People Below Poverty Limit (%)",
       x = "County Name") +
  theme_minimal()
```

```{r, message=FALSE, warning=FALSE, echo = FALSE}
co_county_poverty_plot
```

